The development of children's ability to track and predict turn structure in conversation

### 2016-06-05 paper/CasillasFrank-Anticip.tex
most changes for resubmission

\usepackage{graphicx}
\usepackage{lineno}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{subfig}
\usepackage{tipa}
\usepackage{float}
\usepackage{longtable}
\usepackage[section]{placeins}
\usepackage{rotating}

\linespread{2}
\journal{}

@@

\begin{abstract}
Children begin developing turn-taking skills in infancy but take several years to assimilate their growing knowledge of language into their turn-taking behavior. In two eye-tracking experiments, we measured children's anticipatory gaze to upcoming responders while controlling linguistic cues to turn structure. In Experiment 1, we showed English and non-English conversations to English-speaking adults and children. In Experiment 2, we phonetically controlled lexicosyntactic and prosodic cues in English-only speech. Children spontaneously made anticipatory gaze switches by age two and continued improving through age six. In both experiments, children and adults made more anticipatory switches after hearing questions. Like adults, prosody alone did not improve children's predictive gaze shifts. But, unlike adults, lexical cues alone were not sufficient to improve prediction---children's performance was best overall with lexicosyntax and prosody together. Our findings support an account in which turn prediction emerges in infancy, but then only gradually becomes fully integrated with linguistic processing. 
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}

@@

\subsection{Adult turn taking}

@@

Prior work suggests that adult listeners primarily use lexicosyntactic information to accurately predict upcoming turn structure. De Ruiter and colleagues \citeyearpar{de-ruiter2006} asked participants to listen to snippets of spontaneous conversation and to press a button whenever they anticipated that the current speaker was about to finish his or her turn. The speech snippets were controlled for the amount of linguistic information present; some were normal, but others had flattened pitch, low-pass filtered speech, or further manipulations. With pitch-flattened speech, the timing of participants' button responses was comparable to their timing with the full linguistic signal. But when no lexical information was available, participants' responses were significantly earlier. The authors concluded that lexicosyntactic information\footnote{The ``lexicosyntactic'' condition only included flattened pitch and so was not exclusively lexicosyntactic---the speech would still have residual prosodic structure, including syllable duration and intensity.} was necessary and possibly sufficient for turn-end projection, while intonation was neither necessary nor sufficient. Congruent evidence comes from studies varying the predictability of lexicosyntactic and pragmatic content: adults anticipate turn ends better when they can more accurately predict the exact words that will come next (\citealp{magyari2012}; see also \citealp{magyari2014}). They can also identify speech acts within the first word of an utterance \citep{gisladottir2015}, allowing them to start planning their response at the first moment possible \citep{bogels2015}.

@@

\subsection{Turn taking in development}

@@

The timing of children's responses to their caregivers' speech shows a non-linear pattern. Infants' contingent vocalizations in the first few months of life show very fast timing (though with a lot of vocal overlap). But by nine months, their timing slows down considerably, only to gradually speed up again after 12 months \citep{hilbrink2015}. For children, taking turns with brief transitions between speakers is more difficult than avoiding speaker overlap; children's incidence of overlap is nearly adult-like by nine months, but the timing of their non-overlapped responses stays much longer than the adult 200 msec standard for the next few years \citep{casillas2016, garvey1984, garvey1981, ervin-tripp1979}. This puzzling pattern is likely due to their linguistic development: taking turns on time is easier when the response is a simple vocalization rather than a linguistic utterance. Integrating language into the turn-taking system may therefore be one major factor in children's delayed responses \citep{casillas2016}.

Children, like adults, might use linguistic cues in the ongoing turn to make predictions about upcoming turn structure. Studies of early linguistic development point to a possible early advantage for prosody over lexicosyntax in children's turn-taking predictions. Infants can distinguish their native language's rhythm type from others soon after birth \citep{mehler1988, nazzi2003}; they show preference for the typical stress patterns of their native language over others by 6--9 months (e.g., iambic vs. trochaic), and can use prosodic information to segment the speech stream into smaller chunks from 8 months onward \citep{johnson2001, morgan1995}. Four- to five-month-olds also prefer pauses in speech to be inserted at prosodic boundaries, and by 6 months infants can use prosodic markers to pick out sub-clausal syntactic units, both of which are useful for extracting turn structure from ongoing speech \citep{jusczyk1995, soderstrom2003}. In comparison, children show at best a very limited lexical inventory before their first birthday \citep{bergelson2013, shi2010}.

@@

Keitel and colleagues \citeyearpar{keitel2013} were one of the first to explore how children use linguistic cues to predict upcoming turn structure. They asked 6-, 12-, 24-, and 36-month-old infants, and adult participants to watch short videos of conversation and tracked their eye movements at points of speaker change. They showed their participants two types of conversation videos---one normal and one with flattened pitch (i.e., with flattened intonation contours)---to test the role of intonation in participants' anticipatory predictions about upcoming speech. Comparing children's anticipatory gaze frequency to a random baseline, they found that only 36-month-olds and adults made anticipatory gaze switches more often than expected by chance. Among those, only 36-month-olds were affected by a lack of intonation contours, leading Keitel and colleagues to conclude that children's ability to predict upcoming turn structure relies on their ability to comprehend the stimuli lexicosemantically. They also suggest that intonation might play a secondary role in turn prediction, but only after children acquire more sophisticated, adult-like language comprehension abilities (also see \citealp{keitel2015}).

@@

Second, the conversation stimuli \citet{keitel2013} used were somewhat unusual. The average gap between turns was 900 msec, a duration much longer than typical adult timing, which averages around 200 msec \citep{stivers2009}. The speakers in the videos were also asked to minimize their movements while performing scripted, adult-directed conversation, which would have created a somewhat unnatural interaction. Additionally, to produce more naturalistic conversation, it would have been ideal to localize the sound sources for the two voices in the video (i.e., to have the voices come out of separate left and right speakers). But both voices were recorded and played back on the same audio channel, which may have made it difficult to distinguish the two talkers. Again, we attempt to address these issues in our current study. Despite these minor methodological issues, the Keitel et al. \citeyearpar{keitel2013} study still demonstrates intriguing age-based differences in children's ability to predict upcoming turn structure. Our current work takes this paradigm as a starting point.\footnote{But also see \citet{casillas2012, casillas2013}.}

\subsection{The current study}

Our goal in the current study is to find out when children begin to make predictions about upcoming turn structure and to understand how their predictions are affected by linguistic cues across development. We present two experiments in which we measured children's anticipatory gaze to responders while they watched conversation videos with natural (people speaking English vs. non-English; Experiment 1) and non-natural (puppets with phonetically manipulated speech; Experiment 2) control over the presence of lexical and prosodic cues. We tested children across a wide range of ages (Experiment 1: 3--5 years; Experiment 2: 1--6 years), with adult control participants in each experiment.

We highlight three primary findings: first, although children and adults use linguistic cues to make predictions about upcoming turn structure, they do so primarily in predict speaker transitions after questions (a speech act effect). This intriguing effect, which has not been reported previously, suggests that participants track unfolding speech for cues to upcoming speaker change, which may affect how they use linguistic cues more generally for anticipatory processing in conversation. Second, we find that children make more predictions than expected by chance starting at age two, but that this effect is small at first, and continues to improve through age six. Third, we find no evidence of an early prosody advantage in children's anticipations and, further, no evidence that prosodic or lexical cues alone can substitute for their combination in the full linguistic signal (as is proposed for adults; \citealp{de-ruiter2006}). Instead, anticipation is strongest for stimuli with the full range of cues. In sum, our findings support an account in which turn prediction emerges in infancy, but becomes fully integrated with linguistic processing only gradually across development.

\section{Experiment 1}
\label{sec:exp1}

We recorded participants' eye movements as they watched six short videos of two-person (dyadic) conversation interspersed with attention-getting filler videos. Each conversation video featured an improvised discourse in one of five languages (English, German, Hebrew, Japanese, and Korean). Participants saw two videos in English and one in every other language. The participants, all native English speakers, were only expected to understand the two videos in English. We showed participants non-English videos to limit their access to lexical information while maintaining their access to other cues to turn boundaries (e.g., (non-native) prosody, gaze, inbreaths, phrase final lengthening). Using this method, we compared children and adult's anticipatory looks from the current speaker to the upcoming speaker at points of turn transition in English and non-English videos.

\subsection{Methods}
\label{sec:methods1}

\subsubsection{Participants}

We recruited 74 children between ages 3;0--5;11 and 11 undergraduate adults to participate in the experiment. Our child sample included 19 three-year-olds, 32 four-year-olds, and 23 five-year-olds, all enrolled in a local nursery school. All participants were native English speakers. Approximately one-third (N=25) of the children's parents and teachers reported that their child regularly heard a second (and sometimes third or further) language, but only one child frequently heard a language that was used in our non-English video stimuli, and we excluded his data from analyses. None of the adult participants reported fluency in a second language.

@@

\subsubsection{Materials}

@@

After recording, we combined the audio and video recordings by hand, and cropped each recording to the (approximate) 30-second interval with the most turn activity. Because we recorded the conversations in stereo, the male and female voices came out of separate speakers during video playback. This gave each voice in the videos a localized source (from the left or right loudspeaker). We coded each turn transition in the videos for language condition (English vs. non-English), inter-turn gap duration (in milliseconds), and speech act (question vs. non-question). The non-English stimuli were coded for speech act from a monolingual English-speaker's perspective, i.e., which turns ``sound like'' questions, and which do not: we asked five native American English speakers to listen to the audio recording for each non-English turn and judge whether it sounded like a question. We marked non-English turns as questions when at least 4 of the 5 listeners (80\%) said that the turn ``sounded like a question''.

Because the conversational stimuli were recorded semi-spontaneously, the duration of turn transitions and the number of speaker transitions in each video was variable. We measured the duration of each turn transition from the audio recording associated with each video. We excluded turn transitions longer than 550 msec and shorter than 90 msec, also excluding overlapped transitions, from analysis.\footnote{Overlap occurs when a responder begins a new turn before the current turn is finished. When overlap occurs, observers cannot switch their gaze in anticipation of the response because the response began earlier than expected. Participants expect conversations to proceed with ``one speaker at a time'' \citep{sacks1974}. They would therefore still be fixated on the prior speaker when the overlap started, and would have to switch their gaze \textit{reactively} to the responder.} This left approximately equal numbers of turn transitions available for analysis in the English (N$=$20) and non-English (N$=$16) videos. On average, the inter-turn gaps for English videos (mean$=$318, median$=$302, stdev$=$112 msec) were slightly longer than for non-English videos (mean$=$286, median$=$251, stdev$=$122 msec). The longer gaps in the English videos could give them a slight advantage: our definition of an ``anticipatory gaze shift'' includes shifts that are initiated during the gap between turns (Figure \ref{fig:criterion}), so participants had slightly more time to make anticipatory shifts in the English videos.

@@

\subsubsection{Procedure} 
Participants sat in front of an SMI 120Hz corneal reflection eye-tracker mounted beneath a large flatscreen display. The display and eye-tracker were secured to a table with an ergonomic arm that allowed the experimenter to position the whole apparatus at a comfortable height, approximately 60 cm from the viewer. We placed stereo speakers on the table, to the left and right of the display.

@@

\subsubsection{Data preparation and coding}
\label{sec:algorithm}

@@

We checked each participant's gaze at each turn transition for three characteristics (Figure \ref{fig:criterion}): (1) that the participant fixated on the prior speaker for at least 100 msec at the end of the prior turn, (2) that sometime thereafter the participant switched to fixate on the upcoming speaker for at least 100 ms, and (3) that the switch in gaze was initiated within the first 200 msec of the response turn, or earlier. These criteria guarantee that we only counted gaze shifts when: (1) participants were tracking the previous speaker, (2) switched their gaze to track the upcoming speaker, and (3) did so before they could have simply reacted to the onset of speech in the response. Under this assumption, a gaze shift that was initiated within the first 200 msec of the response (or earlier) was planned \textit{before} the child could react to the onset of speech itself.

As mentioned, most anticipatory switches happen in the inter-turn gap, but we also allowed anticipatory gaze switches that occurred in the final syllables of the prior turn. Early switches are consistent with the distribution of responses in explicit turn-boundary prediction tasks. For example, in a button press task, adult participants anticipate turn ends approximately 200 msec in advance of the turn's end, and anticipatory responses to pitch-flattened stimuli come even earlier \citep{de-ruiter2006}. We therefore allowed switches to occur as early as 200 msec before the end of the prior turn. Again, because it takes 200 msec to plan an eye movement, we counted anticipatory switches, at the latest, 200 msec after the onset of speech. Therefore, for very early and very late switches, our requirement of 100 msec of fixation on each speaker would sometimes extend outside of the transition window boundaries (200 msec before and after the inter-turn gap). The maximally available fixation window was therefore 100 msec before and after the earliest and latest possible switch point (300 msec before and after the inter-turn gap). We did not count switches made during the fixation window as anticipatory. We \textit{did} count switches made during the inter-turn gap. The period of time from the beginning of the possible fixation window on the prior speaker to the end of the possible fixation window on the responder was our total analysis window (300 msec $+$ the inter-turn gap $+$ 300 msec).

@@

\paragraph{Predictions}
We expected participants to show greater anticipation in the English videos than in the non-English videos because of their increased access to linguistic information in English. We also predicted that anticipation would be greater following questions compared to non-questions; questions have early cues to upcoming turn transition (e.g., \textit{wh-} words, subject-auxiliary inversion), and also make a next response immediately relevant. Our third prediction was that anticipatory looks would increase with development, along with children's increased linguistic competence.

\subsection{Results}
\label{sec:results1}

@@

\linespread{1}
\begin{table}[t]
\begin{center}
  \begin{tabular}{llcccc}
    \hline
    Age group & Condition & Speaker & Addressee & Other onscreen & Offscreen\\ 
    \hline
    3 & English & 0.61 & 0.16 & 0.14 & 0.08 \\ 
    4 & English & 0.60 & 0.15 & 0.11 & 0.13 \\ 
    5 & English & 0.57 & 0.15 & 0.16 & 0.12 \\ 
    Adult & English & 0.63 & 0.16 & 0.16 & 0.05 \\ 
    3 & Non-English & 0.38 & 0.17 & 0.20 & 0.25 \\ 
    4 & Non-English & 0.43 & 0.19 & 0.21 & 0.18 \\ 
    5 & Non-English & 0.40 & 0.16 & 0.26 & 0.18 \\ 
    Adult & Non-English & 0.58 & 0.20 & 0.16 & 0.07 \\ 
%   Overall & & 0.49 & 0.17 & 0.18 & 0.16 \\
    \hline
  \end{tabular}
\end{center}
  \caption{Average proportion of gaze to the current speaker and addressee during periods of talk.}
\label{tab:e1_look}
\end{table}
\linespread{2}

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\textwidth]{figures/E1-samples-by-lang-groups-trans-types.png}
\end{center}
\caption{Anticipatory gaze rates across language condition and transition type for the real data (red and blue) and the randomly permuted baselines (gray). Vertical bars represent 95\% confidence intervals.} 
\label{fig:E1-randvsreal}
\end{figure}

>>>>

\usepackage{graphicx}
\usepackage{lineno}
\usepackage{amssymb}
\usepackage{wrapfig}
\usepackage{subfig}
\usepackage{tipa}
\usepackage{float}
\usepackage{longtable}
\usepackage[section]{placeins}
\usepackage{rotating}
%\usepackage{apacite}

%\linespread{2}
\journal{}

@@

\begin{abstract}
Children begin developing turn-taking skills in infancy but take several years to assimilate their growing knowledge of language into their turn-taking behavior. In two eye-tracking experiments, we measured children's anticipatory gaze to upcoming responders while controlling linguistic cues to turn structure. In Experiment 1, we showed English and non-English conversations to English-speaking adults and children. In Experiment 2, we phonetically controlled lexicosyntactic and prosodic cues in English-only speech. Children spontaneously made anticipatory gaze switches by age two and continued improving through age six. In both experiments, children and adults made more anticipatory switches after hearing questions. Consistent with prior findings on adult turn prediction, prosodic information alone did not increase children's anticipatory gaze shifts. But, unlike prior work with adults, lexical information alone was not sufficient either---children's performance was best overall with lexicosyntax and prosody together. Our findings support an account in which turn tracking and prediction emerges in infancy, and then only gradually becomes integrated with linguistic processing.
\end{abstract}

@@

\section*{Introduction}
\label{sec:intro}

@@

\subsection*{Adult turn taking}

@@

Prior work suggests that adult listeners primarily use lexicosyntactic information to accurately predict upcoming turn structure. De Ruiter and colleagues \citeyearpar{de-ruiter2006} asked participants to listen to snippets of spontaneous conversation and to press a button whenever they anticipated that the current speaker was about to finish his or her turn. The speech snippets were controlled for the amount of linguistic information present; some were normal, but others had flattened pitch, low-pass filtered speech, or further manipulations. With pitch-flattened speech, the timing of participants' button responses was comparable to their timing with the full linguistic signal. But when no lexical information was available, participants' responded significantly earlier within the turn. The authors concluded that lexicosyntactic information\footnote{The ``lexicosyntactic'' condition only included flattened pitch and so was not exclusively lexicosyntactic---the speech would still have residual prosodic structure, including syllable duration and intensity.} was necessary and possibly sufficient for turn-end projection, while intonation was neither necessary nor sufficient. Congruent evidence comes from studies varying the predictability of lexicosyntactic and pragmatic content: adults anticipate turn ends better when they can more accurately predict the exact words that will come next (\citealp{magyari2012}; see also \citealp{magyari2014}). They can also identify speech acts within the first word of an utterance \citep{gisladottir2015}, allowing them to start planning their response at the first moment possible \citep{bogels2015}.

@@

\subsection*{Turn taking in development}

@@

The timing of children's responses to their caregivers' speech shows a non-linear pattern. Infants' contingent vocalizations in the first few months of life show very fast timing (though with a lot of vocal overlap). But by nine months, their timing slows down considerably, only to gradually speed up again after 12 months \citep{hilbrink2015}. For children, taking turns with brief transitions between speakers is more difficult than avoiding speaker overlap; children's incidence of overlap is nearly adult-like by nine months, but the timing of their non-overlapped (i.e., gapped) responses remains longer than the adult 200 msec standard for the next few years \citep{casillas2016, garvey1984, garvey1981, ervin-tripp1979}. This puzzling pattern is likely due to children's linguistic development: taking turns on time is easier when their response is a simple vocalization rather than a linguistic utterance. Integrating language into the turn-taking system may therefore be a major factor in children's delayed responses \citep{casillas2016}.

Before children manage to integrate linguistic cues into their turn-taking behaviors (for both turn prediction and production), they can rely on non-verbal interactional cues, including silence (e.g., between turns), eye gaze, body orientation, and gesture, to identify the boundaries of social actions. For example, with little to no linguistic knowledge, children are often able to infer and anticipate desired responses to offers and requests by taking account of their interlocutor's non-verbal communicative behavior, the structure of routine events, and the affordances of the current interactional context \citep{reddy13, nomikou11, shatz78}. With respect to turn taking in particular, children's spontaneous vocalizations during interaction demonstrate a sensitivity to short inter-speaker gaps from infancy \citep{hilbrink2015}, even when they are not yet using language to participate. Thus, before they can anticipate turn structure form linguistic cues, children might instead react to silence as a cue to upcoming speaker change. Interactional silence itself may then serve as an one of children's first cues to turn structure, giving them information about when to respond before they can rely on language.

As children's language competence increases, they can use linguistic cues to make predictions about upcoming turn structure. Studies of early linguistic development point to a possible early advantage for prosody over lexicosyntax in children's turn-taking predictions. Infants can distinguish their native language's rhythm type from others soon after birth \citep{mehler1988, nazzi2003}; they show preference for the typical stress patterns of their native language over others by 6--9 months (e.g., iambic vs. trochaic), and can use prosodic information to segment the speech stream into smaller chunks from 8 months onward \citep{johnson2001, morgan1995}. Four- to five-month-olds also prefer pauses in speech to be inserted at prosodic boundaries, and by 6 months infants can use prosodic markers to pick out sub-clausal syntactic units, both of which are useful for extracting turn structure from ongoing speech \citep{jusczyk1995, soderstrom2003}. In comparison, children show at best a very limited lexical inventory before their first birthday \citep{bergelson2013, shi2010}.

@@

Keitel and colleagues \citeyearpar{keitel2013} were one of the first to explore how children use linguistic cues to predict upcoming turn structure. They asked 6-, 12-, 24-, and 36-month-old infants, and adult participants to watch short videos of conversation and tracked their eye movements at points of speaker change. They showed their participants two types of conversation videos---one normal and one with flattened pitch---to test the role of intonation in participants' anticipatory predictions about upcoming speech. Comparing children's anticipatory gaze frequency to a random baseline, they found that only 36-month-olds and adults made anticipatory gaze switches more often than expected by chance, and only 36-month-olds were affected by a lack of intonation contours. This finding led Keitel and colleagues to conclude that children's ability to predict upcoming turn structure relies on their ability to comprehend the stimuli lexicosemantically. They also suggest that intonation might play a secondary role in turn prediction, but only after children acquire more sophisticated, adult-like language comprehension abilities (also see \citealp{keitel2015}).

@@

Second, the conversation stimuli \citet{keitel2013} used were somewhat unusual. The average gap between turns was 900 msec, a duration much longer than typical adult timing, which averages around 200 msec \citep{stivers2009}. The speakers in the videos were also asked to minimize their movements while performing scripted, adult-directed conversation, which would have created a somewhat unnatural interaction. Additionally, to produce more naturalistic conversation, it would have been ideal to localize the sound sources for the two voices in the video (i.e., to have the voices come out of separate left and right speakers). But both voices were recorded and played back on the same audio channel, which may have made it difficult to distinguish the two talkers. Again, we attempt to address these issues in our current study. Despite these minor methodological issues, the Keitel et al. \citeyearpar{keitel2013} study still demonstrates intriguing age-based differences in children's ability to predict upcoming turn structure. Our current work takes these findings as a starting point.\footnote{But also see \citet{casillas2012, casillas2013}.}

\subsection*{The current study}

Our goal in the current study is to find out when children begin to make predictions about upcoming turn structure and to understand how their predictions are affected by linguistic cues across development. We present two experiments in which we measured children's anticipatory gaze to responders while they watched conversation videos with natural (people speaking English vs. non-English; Experiment 1) and non-natural (puppets with phonetically manipulated speech; Experiment 2) control over the presence of lexical and prosodic cues. We tested children across a wide range of ages (Experiment 1: 3--5 years; Experiment 2: 1--6 years), with adult control participants in each experiment. We additionally tested for the use of one non-verbal cue: inter-turn silence.

We highlight four primary findings: first, although children and adults use linguistic cues to make predictions about upcoming turn structure, they do so primarily to predict speaker transitions after questions (a speech act effect). This intriguing effect, which has not been reported previously, suggests that participants track unfolding speech for cues to upcoming speaker change, which may affect how they use linguistic cues more generally for anticipatory processing in conversation. Second, we find that children make more predictions than expected by chance starting at age two, but that this effect is small at first, and continues to improve through age six, along with children's use of linguistic cues to anticipate answers after question turns. Third, children and adults at all ages often used inter-turn silence (a non-verbal cue to turn structure) to make more predictive gaze switches to the responder, suggesting that non-verbal cues are useful for predicting turn structure early on and continue to be important in adulthood. Finally, we find no evidence for an early prosodic advantage in children's anticipations and, further, no evidence that lexical cues alone are comparable to the full linguistic signal in aiding participants' predictions (as is proposed for adults; \citealp{de-ruiter2006}). Anticipation is strongest for stimuli with the full range of linguistic cues. Our findings support an account in which turn prediction emerges in infancy, but becomes fully integrated with linguistic processing only gradually across development.

\section*{Experiment 1}
\label{sec:exp1}

We recorded participants' eye movements as they watched six short videos of two-person (dyadic) conversation interspersed with attention-getting filler videos. Each conversation video featured an improvised discourse in one of five languages (English, German, Hebrew, Japanese, and Korean). Participants saw two videos in English and one in every other language. The participants, all native English speakers, were only expected to understand the two videos in English. We showed participants non-English videos to limit their access to lexical information while maintaining their access to other cues to turn boundaries (e.g., non-English prosody, gaze, in-breaths, phrase final lengthening). Using this method, we compared children and adult's anticipatory looks from the current speaker to the upcoming speaker at points of turn transition in English and non-English videos.

\subsection*{Methods}
\label{sec:methods1}

\subsubsection*{Participants}

We recruited 74 children ages 3;0--5;11 and 11 undergraduate adults to participate in the experiment. We recruited adult participants through the Stanford University Psychology participant database. Adult participants were either paid or received course credit for their time. Our child sample included 19 three-year-olds, 32 four-year-olds, and 23 five-year-olds, all enrolled in a local nursery school. All participants were native English speakers and volunteered their time. Approximately one-third (N$=$25) of the children's parents and teachers reported that their child regularly heard a second (and sometimes third or further) language, but only one child frequently heard a language that was used in our non-English video stimuli, and we excluded his data from the analyses.\footnote{Multilingual children may make predictions about upcoming turn structure differently from their monolingual peers due to their more varied experiences with language-relevant turn taking cues, but we are unable to test this hypothesis here due to the variability in multilingual language input and languages being learned in our sample. The same note applies to Experiment 2 below.} None of the adult participants reported fluency in a second language.

@@

\subsubsection*{Materials}

@@

After recording, we combined the audio and video recordings by hand, and cropped each recording to the (approximate) 30-second interval with the most turn activity. Because we recorded the conversations in stereo, the male and female voices came out of separate speakers during video playback. This gave each voice in the videos a localized source (from the left or right loudspeaker). We coded each turn transition in the videos for language condition (English vs. non-English), inter-turn gap duration (in milliseconds), and speech act (question vs. non-question). The non-English stimuli were coded for speech act from a monolingual English-speaker's perspective, i.e., which turns ``sound like'' questions, and which do not: we asked five native American English speakers to listen to the audio recording for each non-English turn and judge whether it sounded like a question. We marked non-English turns as questions when at least 4 of the 5 listeners (80\%) said that the turn ``sounded like a question''. This procedure means that the cues for recognizing ``questions'' in the non-English condition only resembled native English question cues, and therefore were likely harder to identify than cues to questionhood in the English condition. However, since participants did not speak the non-English languages and would only therefore ever treat ``question-sounding'' turns as questions, we proceeded with these analyses to see how pervasive question effects were---could they show up even without lexical access in a foreign language? If participants primarily rely on prosodic cues to question turns, it's possible that even non-English prosody can elicit anticipatory gaze switches for question-like turns.

Because the conversational stimuli were recorded semi-spontaneously, the duration of turn transitions and the number of speaker transitions in each video was variable. We measured the duration of each turn transition from the audio recording associated with each video. We excluded turn transitions longer than 550 msec and shorter than 90 msec from analysis, additionally excluding overlapped transitions.\footnote{Overlap occurs when a responder begins a new turn before the current turn is finished. When overlap occurs, observers cannot switch their gaze in anticipation of the response because the response began earlier than expected. Participants expect conversations to proceed with ``one speaker at a time'' \citep{sacks1974}. They would therefore still be fixated on the prior speaker when the overlap started, and would have to switch their gaze \textit{reactively} to the responder.} This left approximately equal numbers of turn transitions available for analysis in the English (N$=$20) and non-English (N$=$16) videos. On average, the inter-turn gaps for English videos (mean$=$318, median$=$302, stdev$=$112 msec) were slightly longer than for non-English videos (mean$=$286, median$=$251, stdev$=$122 msec).

@@

\subsubsection*{Procedure} 
Participants sat in front of an SMI 120Hz corneal reflection eye-tracker mounted beneath a large flatscreen display. The display and eye-tracker were secured to a table with an ergonomic arm that allowed the experimenter to position the whole apparatus at a comfortable height, approximately 60 cm from the viewer. We placed stereo speakers on the table, to the left and right of the display.

@@

\subsubsection*{Data preparation and coding}
\label{sec:algorithm}

@@

We checked each participant's gaze at each turn transition for three characteristics (Figure \ref{fig:criterion}): (1) that the participant fixated on the prior speaker for at least 100 msec at the end of the prior turn, (2) that immediately thereafter the participant switched to fixate on the upcoming speaker for at least 100 ms, and (3) that the switch in gaze was initiated within the first 200 msec of the response turn, or earlier. These criteria guarantee that we only counted gaze shifts when: (1) participants were tracking the previous speaker, (2) switched their gaze to track the upcoming speaker, and (3) did so before they could have simply reacted to the onset of speech in the response. Under this assumption, a gaze shift that was initiated within the first 200 msec of the response (or earlier) was planned \textit{before} the child could react to the onset of speech itself.

As mentioned, most anticipatory switches happen in the inter-turn gap, but we also allowed anticipatory gaze switches that occurred in the final syllables of the prior turn. Early switches are consistent with the distribution of responses in explicit turn-boundary prediction tasks. For example, in a button press task, adult participants anticipated turn ends approximately 200 msec in advance of the turn's end, and anticipatory responses to pitch-flattened stimuli came even earlier \citep{de-ruiter2006}. We therefore allowed switches to occur as early as 200 msec before the end of the prior turn. Again, because it takes 200 msec to plan an eye movement, we counted anticipatory switches, at the latest, 200 msec after the onset of speech. Therefore, for very early and very late switches, our requirement of 100 msec of fixation on each speaker would sometimes extend outside of the transition window boundaries (200 msec before and after the inter-turn gap). The maximally available fixation window was therefore 100 msec before and after the earliest and latest possible switch point (300 msec before and after the inter-turn gap). We did not count switches made during the fixation window as anticipatory. We \textit{did} count switches made during the inter-turn gap. The period of time from the beginning of the possible fixation window on the prior speaker to the end of the possible fixation window on the responder was our total analysis window (300 msec $+$ the inter-turn gap $+$ 300 msec).

@@

\paragraph{Predictions}
We expected participants to show greater anticipation in the English videos than in the non-English videos because of their increased access to linguistic information in English. We also predicted that anticipation would be greater following questions compared to non-questions; questions have early cues to upcoming turn transition (e.g., \textit{wh-} words, subject-auxiliary inversion), and also make a next response immediately relevant. Our third prediction was that anticipatory looks would increase with development, along with children's increased linguistic competence. Finally, we predicted that transitions with longer inter-turn gaps would show greater anticipation because longer gaps provide (a) more time to make a gaze switch and (b) are themselves a cue to possible upcoming speaker switch.

\subsection*{Results}
\label{sec:results1}

@@

\linespread{1}
\begin{table}[t]
\begin{center}
  \begin{tabular}{llcccc}
    \hline
    Age group & Condition & Speaker & Addressee & Other onscreen & Offscreen\\ 
    \hline
    3 & English & 0.61 & 0.16 & 0.14 & 0.08 \\ 
    4 & English & 0.60 & 0.15 & 0.11 & 0.13 \\ 
    5 & English & 0.57 & 0.15 & 0.16 & 0.12 \\ 
    Adult & English & 0.63 & 0.16 & 0.16 & 0.05 \\ 
    3 & Non-English & 0.38 & 0.17 & 0.20 & 0.25 \\ 
    4 & Non-English & 0.43 & 0.19 & 0.21 & 0.18 \\ 
    5 & Non-English & 0.40 & 0.16 & 0.26 & 0.18 \\ 
    Adult & Non-English & 0.58 & 0.20 & 0.16 & 0.07 \\ 
%   Overall & & 0.49 & 0.17 & 0.18 & 0.16 \\
    \hline
  \end{tabular}
\end{center}
  \caption{Average proportion of gaze to the current speaker and addressee during periods of talk.}
\label{tab:e1_look}
\end{table}
%\linespread{2}

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\textwidth]{figures/E1-samples-by-lang-groups-trans-types.png}
\end{center}
\caption{Anticipatory gaze rates across language condition and transition type for the real and randomly permuted datasets. Vertical bars represent 95\% confidence intervals.} 
\label{fig:E1-randvsreal}
\end{figure}

\linespread{1}
\begin{table}[h!]
\begin{small}
\begin{center}
  \begin{tabular}{lcccl}
  \textbf{\textit{Children}} &&&& \\
    \hline
			           &  Estimate & Std. Error & \textit{z} value & Pr($>$$|$\textit{z}$|$) \\    
    \hline
    (Intercept)   													& -0.604	& 1.242	& -0.486  & 0.627		\\
    Age             												& -0.002	& 0.261	& -0.009  & 0.993		\\
    LgCond$=$\textit{non-English} 						& -3.65		& 1.16	& -3.146  & 0.002 **	\\
    Type$=$\textit{non-Question}      					& -2.95		& 1.13	& -2.61	  & 0.009 **	\\
    GapDuration      											&  2.247	& 3.194	&  0.704  & 0.482		\\
    Age*LgCond$=$\textit{non-English} 					&  0.5		& 0.212	&  2.353  & 0.019 *	\\
    Age*Type$=$\textit{non-Question} 					&  0.009	& 0.196	&  0.044  & 0.965		\\
    LgCond$=$\textit{non-English}*						&  2.692	& 1.347	&  1.999  & 0.046 *	\\
    \hspace*{5mm} Type$=$\textit{non-Question} &&&& \\
    Age*GapDuration												& -0.577	& 0.627	& -0.921  & 0.357		\\
    LgCond$=$\textit{non-English}*GapDuration		&  1.143	& 2.287	&  0.5	  & 0.617		\\
    Type$=$\textit{non-Question}*GapDuration			&  5.519	& 2.282	&  2.418  & 0.016 *	\\
    Age*LgCond$=$\textit{non-English}*					& -0.433	& 0.304	& -1.426  & 0.154		\\
    \hspace*{5mm} Type$=$\textit{non-Question} &&&& \\
    \hline
  &&&& \\
  \textbf{\textit{Adults}} &&&& \\
    \hline
           &  Estimate & Std. Error & \textit{z} value & Pr($>$$|$\textit{z}$|$) \\    
    \hline
    (Intercept)														& -0.584   & 0.64  & -0.913 & 0.361			\\
    LgCond$=$\textit{non-English}							& -0.059   & 0.751 & -0.079 & 0.937			\\
    Type$=$\textit{non-Question}							& -3.298   & 0.933 & -3.536 & 0.0004 ***	\\
    GapDuration														&  0.132   & 1.766 &  0.075 & 0.941			\\ 
    LgCond$=$\textit{non-English}*						&  1.234   & 0.629 &  1.961 & 0.0498 *		\\
    \hspace*{5mm} Type$=$\textit{non-Question} &&&& \\
    LgCond$=$\textit{non-English}*GapDuration		& -1.519   & 2.192 & -0.693 & 0.488			\\
    Type$=$\textit{non-Question}*GapDuration			&  7.116   & 2.195 &  3.241 & 0.001 **		\\
    \hline
  \end{tabular}
\end{center}
  \end{small}
  \caption{Model output for children and adults' anticipatory gaze switches in Experiment 1.}
\label{tab:E1-models}
\end{table}
%\linespread{2}

\subsubsection*{Statistical models}
\label{sec:models1}
